---
title: "Tidytuesday_extinct_plants"
author: "Jone"
date: "17/8/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages, warning=FALSE}
library(tidyverse)
library(tidytuesdayR)
# map
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
# combine figures
library(patchwork)
library(cowplot)
```

```{r}
datos <- tt_load("2020-08-18")
```
Let's explore the data:

```{r}
plants <- datos$plants
```


We have 500 diferent plants, and 24 variables: Name and taxonomic group (6 groups with very different n-s, the majority is from the same group), Country and continent of origin; Year last seen (we can see a period with start and end year e.g. 2000-2020, or only end year e.g. before 2020); 12 threats, 6 actions and extinction category (e.g. Extinct or Extinct in the wild).

Let's tidy the year column, I want a column with the last year.

I also want two more columns: total_threats and total actions (I am not counting the NAs, so they are going to be total known threats and actions).

```{r}
plants <- plants %>%
  mutate(
    last_year = str_extract(
      year_last_seen, "\\d{4}$"),
    total_threats = rowSums(.[6:16]),
    total_actions = rowSums(.[18:22]))

```



```{r}
plants_continent <- plants %>%
  group_by(continent) %>%
  summarise(
    extinctions = n(),
    sum_threats = sum(total_threats),
    sum_actions = sum(total_actions)
  )

plants_country <- plants %>%
  group_by(country) %>%
  summarise(
    extinctions = n(),
    sum_threats = sum(total_threats),
    sum_actions = sum(total_actions),
    continent = first(continent)
    ) %>%
 mutate(
    country = case_when(
      country == "Cabo Verde" ~ "Cape Verde",
      country == "Congo"      ~ "Democratic Republic of the Congo",
      country == "Pitcairn"   ~ "Pitcairn Islands",
      country == "Saint Helena, Ascension and Tristan da Cunha" ~ "Saint Helena",
      country == "Sao Tome and Principe" ~ "São Tomé and Principe",
      country == "Viet Nam"              ~ "Vietnam",
      TRUE                      ~  country
    )
  )
```

There must be a better way to do that (mutate - case_when)... 

## Making the visualization


We are going to use `rnaturalearth::ne_countries()` to pull country data and sf classes.

```{r}
world <- ne_countries(scale = "medium", returnclass = "sf")
```

And we are going to join the data with plants data by country in order to do the map.

```{r}
map_plants_data <- left_join(
  world, plants_country, 
  by = c("name_long" = "country"))
```



```{r}
map_plants_data <- map_plants_data %>%
  mutate(
    extinctions = case_when(
      is.na(extinctions) ~ "zero",
      TRUE ~ as.character(extinctions)),
    extinctions = case_when(
      extinctions == "zero" ~ 0,
      TRUE ~ as.numeric(extinctions)),
    a_cut = cut(extinctions,
                breaks=c(-Inf, 0, 2, 4, 6, 8, 10, Inf),
                labels=c("0", "1-2", "3-4", "5-6", "7-8", "9-10", "+10"))
  ) 
```


```{r}
map_plant <- ggplot(map_plants_data) +
  geom_sf(aes(fill=a_cut)) +
  coord_sf(crs = "+proj=wink1") +
  scale_fill_manual(values = c("#f0f9e2", "#d6a6b1", "#ad4e64", "#99223e", "#6b172b", "#5b1425", "#000000"),
                    name = "# of extinct plants") +
  guides(fill = guide_legend(title.position = "top",
                             title.hjust = 0.5, nrow = 1,
                             label.position = "bottom",
                             keyheight = 0.8,
                             keywidth = 0.8
                             )) +
  theme_minimal() +
  theme(legend.position = c(0.50, -0.15),
        legend.title = element_text(size = 9),
        legend.text = element_text(size = 7),
        panel.grid.major = 
          element_line(color = gray(.5),
                       linetype = "dashed", size = 0.5))

```



```{r}
cols <- c("Madagascar" = "red4", "United States" = "red4", "Ecuador" = "red4", "Tanzania" = "salmon4", "Malaysia" = "lightsalmon", "Burundi" = "lightsalmon", "Guinea" = "lightsalmon", "Indonesia" = "lightsalmon", "South Africa" = "lightsalmon", "New Caledonia" = "lightsalmon", "Saint Helena" = "lightsalmon", "Brazil" = "lightsalmon")

extinctions_bar <- plants_country %>%
  top_n(12, extinctions) %>%
  group_by(country) %>%
  summarise(extinctions = sum(extinctions)) %>%
  ggplot(aes(y = reorder(country, extinctions),
             x = extinctions)) +
  geom_col(aes(fill = country), col = "black") +
  scale_x_continuous(limits = c(0,120)) +
  scale_fill_manual(values = cols) +
  geom_text(aes(label = extinctions), hjust = -0.25) +
  theme_minimal() +
  theme(legend.position = "none",
        panel.grid = element_blank(),
        axis.title = element_blank(),
        axis.line.x = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_text(color = "black")
        )
  
  
```

```{r}
title2 <- ggplot(data.frame(x = 1:20, y = 1:5)) +
    labs(
    title = "Number of extinct plants by country of origin",
    subtitle = "500 plant species are considered extinct as of 2020. The majority of them were endemic to Madgascar,
    United States (Hawaiian islands) and Ecuador.",
    caption = "Data: IUCN | Viz: @jonealiri"
    ) +
  theme(plot.caption = element_text(hjust = 0))
```



```{r}
title2 + map_plant + extinctions_bar +
  plot_layout(widths = c(0,5,2))
```

```{r}
ggsave("extinct_plants.pdf",
  width = 10, height = 5.5, device = cairo_pdf)

pdftools::pdf_convert("extinct_plants.pdf",
  format = "png", dpi = 350)
```



